---
title: "data_analysis"
author: "finn krauss"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```



## Load Packages

```{r}
# installs packages if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(bcp, changepoint, ggrepel,
  ggplot2, tidyverse, FSA, strucchange,
  MASS, segmented)
```



## Import data

```{r}
load("data/Herring_full.Rdata")

source("R/functions.R")
```



# Change-points
**Analysis:**
`bcp()` is MCMC-based; `changepoint_analysis()` fixes the seed and runs 5000
iterations so the changepoint set is reproducible (verified stable across seeds).

```{r}
results_cpt <- changepoint_analysis(Herring_full,
                                          region_name = "Full Stock",
                                          year_col = "year",
                                          Q = 6,
                                          bcp_threshold = 0.5,
                                          ssb_col = "SSB")
```


**Results:**
```
CPT changepoint years:  1950, 1964, 1966, 1986, 2000, 2018
BCP changepoint years:  1966, 1983, 2000

Consensus changepoints (± 1 yr, min 5 yr apart): 1966, 2000
CPT only: 1950, 1964, 1986, 2018
BCP only: 1983
```


**Plot:**

```{r}
plot_SSB_cpt_full <- plot_SSB_cpt(data = Herring_full,
              changepoints = c(1966,2000),
              ribbon_colors = c("#201124", "#3e478d", "#318ca5"),
              vline_colors = c("black", "black"),
              component = "Full Stock",
              SSB_column = "SSB",
              l_bnd_column = "l_bnd",
              u_bnd_column = "u_bnd",
              show_hlines = TRUE)
plot_SSB_cpt_full
```



**Save plots**

```{r}
ggsave(filename = file.path("plots/", "SSB_cpt_full.png"),
       plot = plot_SSB_cpt_full,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)
```



# Hysteresis
**Analysis:**
The SSB-F cross-correlation (complete observations only) indicates a **3-year
lag**; `breakpoint_analysis()` computes the lagged SSB internally, and a lagged
column is also added for the hysteresis plot.

```{r}
ok <- complete.cases(Herring_full[c("SSB", "F")])
ccf(Herring_full$SSB[ok], Herring_full$F[ok], type = "correlation")

Herring_full$SSB_lag <- dplyr::lag(Herring_full$SSB, 3)

results_brkpt <- breakpoint_analysis(Herring_full, "SSB", "F", "year",
                                    lag_years = 3,
                                    region_name = "Full Stock")
```


**Results:**
```
Optimal number of breaks: 4
Best breakpoint F values: 0.25, 1.43, 0.43, 0.22
Best breakpoint years:    1960, 1971, 1986, 2000
```

**Plot:**

```{r}
plot_hysteresis_full <- plot_hysteresis(Herring_full, results_brkpt$break_years, "Full Stock")
plot_hysteresis_full
```


**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "hysteresis_full.png"),
       plot = plot_hysteresis_full,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)
```



# SRR Non-Stationary
**Analysis:**
The SRR models pair R(y) with SSB(y) (same-year, no larval-to-recruitment lag),
following Blöcker et al. 2023.

```{r}
source("R/model_comp_functions.R")

models_SRR <- run_srr_analysis(
  data_for_srr = Herring_full,
  recruit_col = "Recruitment",
  ssb_col = "SSB")

print(models_SRR$comparison_table)
```


RMSE and R² are computed on the response scale for all models; AIC is only
reported for models fitted on the raw response (log-scale likelihoods are not
comparable). **strucchange has the lowest in-sample RMSE of all models.**

```{r}
### Prep data for LOOCV & run loocv for all models
dat <- Herring_full %>%
  mutate(R_log = log(Recruitment),
         SSB_log = log(SSB)) %>%
  dplyr::select(year, Recruitment, SSB, R_log, SSB_log) %>%
  na.omit() %>%
  setNames(c("Year", "R", "SSB", "R_log", "SSB_log")) %>%
  as.data.frame()

rmse_mean_herring <- cross_valid(dataset = dat)
```


**LOOCV Results:** (all 78 folds contribute for every model — see
`rmse_mean_herring$n_test_folds`):

```
        linear    beverton    ricker      segmented   seg.log     segm.glm    strucchange
TEST:   10164856  9080981     10040655    9437569     9475234     9394547     12535650
TRAIN:  12568771  12158586    12366932    11012203    11184582    11014416    8786426
```

strucchange has the lowest train RMSE but the highest test RMSE of all models,
it fits the observed series closely while predicting held-out years poorly. The
stationary SRR models predict better but describe the series worse, consistent
with strucchange capturing regime structure rather than a single predictive
stock-recruitment relationship.

```{r}
results_SRR <- srr_breakpoint_analysis(Herring_full, "SSB", "Recruitment", "year", "Herring Full")
```


**Results:**

```
Optimal number of breaks: 2
Best breakpoint SSB values: 264793.8, 2391903.7
Best breakpoint years:      1981, 2002
```

**Plot:**

```{r}
plot_SRR_full <- plot_SRR(Herring_full, break_years = results_SRR$break_years, "Full Stock", "strucchange")
plot_SRR_full
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "SRR_full.png"),
       plot = plot_SRR_full,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)
```