---
title: "data_analysis_component"
author: "finn krauss"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```



## Load Packages

```{r}
# installs packages if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(bcp, changepoint, ggrepel,
  ggplot2, tidyverse, FSA, strucchange,
  MASS, segmented, gridExtra, here)
```



## Import data & component configuration
The `components` list holds the per-component configuration:

* **name** - component name in `Herring_LAI`
* **label** - label used in plot titles and printed output
* **lag** - SSB-F time lag in years, chosen from the CCF plots below
* **srr_nudge** - label nudges for the SRR plots
  (first year, breakpoint, last year, more breakpoints)

In all changepoint plots the CI ribbons are colored by the **full-stock phases**
(`full_changepoints`: pre-1966 #201124, 1966–1983 #3e478d, 1983–2000 #318ca5,
post-2000 #5ccdaa), so the full-stock changepoints are visible in every
component panel and it is directly comparable how the components' own shifts
align with them. Each component's own consensus changepoints are drawn as thin
black lines on top. All four components now have consensus changepoints (see
methodology note below) - the "Buchan and Downs have no changepoints" split
reported previously does not hold up.

**Methodology note:** `changepoint_analysis()`'s BCP step was reworked twice
this update. `bcp()` is now called with its published defaults (mcmc=500,
burnin=50, no forced seed), matching the original inspiration this project is
based on (Gutte et al.,
[tipping_northsea_fish](https://github.com/HeleneGutte/tipping_northsea_fish),
`1_Change_points_SSB.R`). Two problems make a *single* such call unreliable
on this data, both confirmed rather than assumed:

1. A single unseeded run is a noisy point estimate - the reference repo's own
   comment notes re-running changed its changepoint set.
2. Raw SSB magnitude (order 1e5-1e6) can trigger a floating-point underflow in
   `bcp()`'s C++ likelihood computation that silently zeroes the *entire*
   posterior - confirmed against the package source (catastrophic cancellation
   in a naive one-pass sum-of-squares term) and against a diagnostic no
   genuine "no evidence" result ever produces (white-noise references give
   small but graded, never-exactly-zero probabilities throughout; the
   affected series here return bit-identical 0 at every single year).

Neither is fixable by picking one lucky seed or one lucky rescaling - both
were tested extensively (including against a broader panel of transforms and
against seed sweeps of 100 runs) and shown to be case-by-case unpredictable:
the same rescaling that recovers one series can silently re-break a different
one that was already fine raw. What *is* reliable, and what the function now
does: run `bcp()` across 100 seeds per series and report, per year, the
fraction of seeds whose posterior probability clears 0.5 (falling back to a
numerically-safer rescaling, chosen from a small fixed panel, only when raw is
degenerate for every seed). A year is "BCP-supported" when that fraction meets
the region's `bcp_threshold`, not when one seed's posterior happened to.

This resolved both an earlier false negative (Buchan and Downs were never
really changepoint-free - both were hitting the underflow already before this
data update, confirmed by rerunning the exact pre-update committed data) and a
false negative introduced mid-investigation (a single-pinned-seed rescaled fix
initially looked like it made the full-stock 2000 changepoint disappear with
the 2026 data; the 100-seed sweep shows 2000 is robustly supported in both the
old and new data at comparable confidence - that apparent loss was an artifact
of trusting one seed's threshold crossing).

```{r}
load("data/Herring_LAI.Rdata")

source("R/functions.R")

### Full-stock phases used to color the CI ribbons in all changepoint plots
full_changepoints <- c(1966, 1983, 2000)
full_ribbon <- c("#201124", "#3e478d", "#318ca5", "#5ccdaa")
full_vline <- c("black", "black", "black")

components <- list(
  SHOR = list(
    name = "Shetland_Orkney",
    label = "Shetland / Orkney",
    lag = 1,
    srr_nudge = list(
      list(nudge_y = 9, nudge_x = -0.15),
      list(nudge_y = 0, nudge_x = -0.2),
      list(nudge_y = 20, nudge_x = 0),
      list(nudge_y = -0, nudge_x = 0.2))),
  Buchan = list(
    name = "Buchan",
    label = "Buchan",
    lag = 3,
    srr_nudge = list(
      list(nudge_y = 5, nudge_x = 0),
      list(nudge_y = 5, nudge_x = 0),
      list(nudge_y = 25, nudge_x = 0),
      list(nudge_y = 10, nudge_x = -0.1))),
  Banks = list(
    name = "Banks",
    label = "Banks",
    lag = 5,
    srr_nudge = list(
      list(nudge_y = 10, nudge_x = 0),
      list(nudge_y = 8, nudge_x = 0),
      list(nudge_y = 15, nudge_x = 0),
      list(nudge_y = 0, nudge_x = 0.05))),
  Downs = list(
    name = "Downs",
    label = "Downs",
    lag = 1,
    srr_nudge = list(
      list(nudge_y = 7, nudge_x = 0),
      list(nudge_y = 4, nudge_x = 0),
      list(nudge_y = 7, nudge_x = 0),
      list(nudge_y = 5, nudge_x = 0))))

### Split data into components
Herring <- lapply(components, function(cfg) {
  Herring_LAI[Herring_LAI$component == cfg$name, ]
})

rm(Herring_LAI)

### Helper to strip axis titles when stacking plots into grids
strip_axes <- function(p) {
  p + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
}
```



# Change-Points
**Analysis:**
`bcp()` estimates posterior probabilities by MCMC and is called here with its
published defaults (mcmc=500, burnin=50), matching the original inspiration
(Gutte et al.,
[tipping_northsea_fish](https://github.com/HeleneGutte/tipping_northsea_fish),
`1_Change_points_SSB.R`). A single such run is a noisy point estimate - that
repo's own comment notes re-running changed its changepoint set, which is why
an earlier version of this pipeline pinned a seed and ran 5000 iterations
instead. That "fix" only made one specific (possibly wrong) outcome
reproducible; it didn't address the actual instability, and separately
obscured a numerical underflow bug (see below). `changepoint_analysis()` now
runs `bcp()` across 100 seeds per series and reports, per year, the fraction
of seeds whose posterior probability clears 0.5 - "BCP-supported" means
supported in at least `bcp_threshold` of the seeds, not that one seed's
posterior happened to cross it.

```{r}
### Q is a real search cap (see note above), so it should scale with series
### length to keep changepoint density comparable across analyses: full stock
### uses Q = 6 over 79 years (1947-2025); components span 53 years
### (1972-2024), so Q = 6 * 53/79 ~= 4 keeps the same points-per-decade rate
### instead of requesting the full-stock count from a much shorter series.
component_Q <- round(6 * 53 / 79)

results_cpt <- list()
for (key in names(components)) {
  results_cpt[[key]] <- changepoint_analysis(Herring[[key]],
                                             region_name = components[[key]]$label,
                                             year_col = "year",
                                             Q = component_Q,
                                             bcp_threshold = 0.5,
                                             ssb_col = "SSB_component")
}
```


**Results:**
**Shetland / Orkney**
```
CPT changepoint years:  1983, 1990, 1994, 2005
BCP support (100 seeds): 1983=100%, 1990=100%, 1994=100%, 1996=91%, 1997=52%, 2002=100%, 2005=100%, 2014=100%
Consensus changepoints (± 1 yr, min 5 yr apart): 1983, 1990, 2005
```

**Banks**
```
CPT changepoint years:  1982, 1990, 2000, 2005
BCP support (100 seeds): 1983=100%, 1990=100%, 2000=100%, 2001=100%, 2005=100%, 2009=98%, 2019=83%
Consensus changepoints (± 1 yr, min 5 yr apart): 1982, 1990, 2000, 2005
```

**Buchan** (previously mis-reported here as "no changepoints" - a numerical
underflow bug in `bcp()`, not a real absence of structure; confirmed present
in the pre-2026-update data too, so this was never actually caused by the data
update)
```
CPT changepoint years:  1986, 1993, 2009, 2018
BCP support (100 seeds): 1989=100%, 1990=100%, 1991=100%, 1992=100%, 1994=100%, 2001=100%, 2010=100%, 2016=84%, 2017=100%, 2018=99%
Consensus changepoints (± 1 yr, min 5 yr apart): 1993, 2009, 2018
```
Buchan has strong, robust BCP support clustering across 1989-1994, 2001,
2010-2011 and 2016-2018. At the reduced (length-matched) `Q`, CPT's nearest
candidate to that first cluster is 1993 rather than 1989 - the underlying BCP
evidence for the cluster is unchanged, only which specific nearby year
represents it in the consensus shifts, since fewer CPT candidates are
requested. Buchan is no longer in the "no changepoints" group.

**Downs** (also previously mis-reported as "no changepoints" for the same
underflow reason)
```
CPT changepoint years:  1987, 2001, 2011, 2021
BCP support (100 seeds): 2001=100%, 2011=100%, 2015=99%, 2021=100%
Consensus changepoints (± 1 yr, min 5 yr apart): 2001, 2011, 2021
```
Downs has strong, robust BCP support at 2001, 2011 and (with the extra year of
data in this update) 2021. Downs is no longer in the "no changepoints" group.

*Interpretation:* all four components now have BCP-supported consensus
changepoints - the "Buchan and Downs have no changepoints" contrast reported
previously never reflected reality; both were hitting the same `bcp()`
underflow bug, already present before this data update, not something the
2026 revision caused. For Downs, its SSB series shows consensus changepoints
in time (2001, 2011, 2021), while its stock-recruitment relationship (SSB vs
Recruitment, see SRR analysis below) shows **no** breakpoint at all - i.e. the
*level* of SSB shifted at those years, but the functional relationship between
SSB and the recruitment it produces did not. For Buchan the SSB changepoints
(1993, 2009, 2018) and the SRR breakpoint structure (see below) are broadly
consistent, both point to periodic regime shifts.



## Phase timeline: full stock vs. components

**Method:** `phase_summary()` fits an OLS regression (value ~ year) on every
point within each phase (not just its endpoints) and reports the slope as an
annualised relative growth rate (%/year, normalized by the phase's own mean so
phases and series of very different absolute scale are comparable). Where a
reference series is supplied it also reports the phase's mean level as a share
of that reference (here: component SSB as % of full-stock SSB in the same
years) and how that share moved from the component's own previous phase - two
phases can have the same slope while one is gaining relative weight in the
total stock and the other is losing it, which the slope alone cannot show.

Component data only starts in 1972 (`Herring_LAI`), so comparisons below only
use the time frame where component data actually exists - the full stock's
own 1947-1966 phase has no component-level counterpart at all, and the
1966-1983 phase is only covered by components from 1972 onward.

For Buchan, 1983 is included as an extra boundary alongside its formally
detected consensus changepoints (1993, 2009, 2018): a rise visible in the SSB
plot right where the full stock itself turns from collapse to recovery, which
BCP/CPT did not separately detect as its own changepoint (it falls inside the
long 1972-1993 span). Splitting it out this way is an informal, visually
motivated read for interpretation - not a claim that a fourth statistical
changepoint exists - but it changes the story enough (see below) to be worth
showing explicitly rather than folding into one long phase.

```{r}
### Loaded here (rather than only later, at the plot-grid step) because this
### phase-timeline comparison needs it first
load("data/Herring_full.Rdata")

full_ssb <- Herring_full %>% dplyr::filter(year <= 2025) %>% dplyr::select(year, SSB)

phase_tables <- list(
  `Full Stock` = phase_summary(Herring_full %>% dplyr::filter(year <= 2025), "SSB", "year",
                               c(1947, 1966, 1983, 2000, 2025), "Full Stock"),
  `Shetland / Orkney` = phase_summary(Herring[["SHOR"]], "SSB_component", "year",
                                      c(range(Herring[["SHOR"]]$year), 1983, 1990, 2005), "Shetland / Orkney",
                                      scale_data = full_ssb, scale_value_col = "SSB"),
  Banks = phase_summary(Herring[["Banks"]], "SSB_component", "year",
                        c(range(Herring[["Banks"]]$year), 1982, 1990, 2000, 2005), "Banks",
                        scale_data = full_ssb, scale_value_col = "SSB"),
  Buchan = phase_summary(Herring[["Buchan"]], "SSB_component", "year",
                         c(range(Herring[["Buchan"]]$year), 1983, 1993, 2009, 2018), "Buchan (1983 indicated)",
                         scale_data = full_ssb, scale_value_col = "SSB"),
  Downs = phase_summary(Herring[["Downs"]], "SSB_component", "year",
                        c(range(Herring[["Downs"]]$year), 2001, 2011, 2021), "Downs",
                        scale_data = full_ssb, scale_value_col = "SSB"))

for (tbl in phase_tables) print(tbl, row.names = FALSE)
```

**Reading the table:**

* **Full stock:** decline (1947-66, sig.) into the collapse (1966-83, sig.),
  a recovery (1983-2000) and a recent flat-to-declining phase (2000-25) that
  are directionally clear but don't individually clear p<0.05 on their own -
  18-26 points against high year-to-year noise isn't enough to confirm either
  on its own merits.
* **Buchan, with 1983 treated as real:** already rising fast before 1983
  (+32%/yr, small base, 7.5% share), then a second strong rise (+18%/yr)
  takes its share of the total stock from 7.5% to 40.6% at exactly the point
  the full stock itself turns from collapse to recovery - coincident with the
  full stock, not a decade-late lag as the single-phase (1972-1993) view
  suggested. The share then nearly round-trips back down to 15.5% through
  1993-2009 despite a merely "flat" slope (the rest of the stock outgrew
  Buchan, not that Buchan shrank), before a second, smaller surge-and-fade
  cycle (2009-18 up, 2018-24 down). Buchan moves in surge-and-fade cycles
  rather than tracking one sustained trend.
* **Downs:** a single uninterrupted, significant rise (1972-2001) spans both
  of the full stock's early regimes with no kink at 1983 - Downs simply did
  not participate in the collapse/recovery cycle the way the others did.
  From 2001 its own slope loses significance in two of the three following
  phases even as its share of the total stock keeps climbing in jumps (33% to
  30% to 45%) rather than a continued smooth climb - "unstable" describes the
  *path*, not the *direction*, which never reverses. Downs has gone from ~8%
  to ~45% of total stock SSB and may now be the largest single component.
* **SHOR and Banks** both show a "slice grew, pie grew faster" pattern during
  1983-2000: SHOR's own SSB rose (+8.8%/yr) while its share of the total fell
  (64% to 41%), because it held an outsized share of a very small total
  during the collapse trough and the recovery was carried disproportionately
  by others. Banks' share fell to 6% by 2000 for the same reason, despite a
  near-flat own-slope.
* **None of the four components leads the full stock in timing** - Buchan's
  first real shift is coincident (not preceding) the full stock's 1983 turn;
  the closest thing to a lead anywhere (Banks, 1982 vs. 1983) is one year,
  within noise.

**Plot:**

```{r}
plots_SSB_cpt <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  cpts <- results_cpt[[key]]$consensus$changepoint_years
  plots_SSB_cpt[[key]] <- plot_SSB_cpt(data = Herring[[key]],
                changepoints = cpts,
                ribbon_changepoints = full_changepoints,
                ribbon_colors = full_ribbon,
                vline_colors = rep("black", length(cpts)),
                component = cfg$label,
                SSB_column = "SSB_component",
                l_bnd_column = "l_bnd_component",
                u_bnd_column = "u_bnd_component",
                show_hlines = FALSE)
  print(plots_SSB_cpt[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "SSB_cpt_components.png"),
       plot = grid.arrange(grobs = lapply(plots_SSB_cpt, strip_axes),
                           ncol = 2,
                           bottom = "Year",
                           left = "SSB (million t)"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```

**Plot Grid:**

```{r}
load("data/Herring_full.Rdata")

### Align the SSB year axis with the F and R columns (fixed 1947-2024 range),
### drop the per-panel x-axis title (one shared "Year" label sits below the
### grid), and drop the per-panel plot title (the component name is already
### the row label - the F and R columns never had one of these to begin with)
align_years <- function(p) {
  suppressWarnings(p + xlim(1947, 2024) +
                      theme(axis.title.x = element_blank(), plot.title = element_blank()))
}

plot_SSB_cpt_full <- plot_SSB_cpt(data = Herring_full,
                                  changepoints = full_changepoints,
                                  ribbon_colors = full_ribbon,
                                  vline_colors = full_vline,
                                  component = "Full Stock",
                                  SSB_column = "SSB",
                                  l_bnd_column = "l_bnd",
                                  u_bnd_column = "u_bnd",
                                  show_hlines = TRUE)

plot_F_full <- plot_F_cpt(data = Herring_full,
                          changepoints = full_changepoints,
                          ribbon_colors = full_ribbon,
                          vline_colors = full_vline,
                          component = "Full Stock",
                          show_hlines = TRUE)

plot_R_full <- plot_Recruitment_cpt(data = Herring_full,
                                    changepoints = full_changepoints,
                                    ribbon_colors = full_ribbon,
                                    vline_colors = full_vline,
                                    component = "Full Stock")

# Component F and R plots (ribbons colored by the full-stock phases)
plots_F <- list()
plots_R <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  cpts <- results_cpt[[key]]$consensus$changepoint_years

  plots_F[[key]] <- plot_F_cpt(data = Herring[[key]],
                               changepoints = cpts,
                               ribbon_changepoints = full_changepoints,
                               ribbon_colors = full_ribbon,
                               vline_colors = rep("black", length(cpts)),
                               component = cfg$label,
                               show_hlines = TRUE)

  plots_R[[key]] <- plot_Recruitment_cpt(data = Herring[[key]],
                                         changepoints = cpts,
                                         ribbon_changepoints = full_changepoints,
                                         ribbon_colors = full_ribbon,
                                         vline_colors = rep("black", length(cpts)),
                                         component = cfg$label)
}

# Assemble the grid with row labels and column headers
library(grid)

make_row_label <- function(label) {
  textGrob(label, rot = 90, gp = gpar(fontsize = 12, fontface = "bold"))
}
make_col_header <- function(label) {
  textGrob(label, gp = gpar(fontsize = 12, fontface = "bold"))
}

grid_grobs <- c(
  # Header row
  list(nullGrob(),
       make_col_header("Fishing Mortality"),
       make_col_header("SSB with Changepoints"),
       make_col_header("Recruitment")),
  # Full stock row
  list(make_row_label("Full Stock"), plot_F_full,
       align_years(plot_SSB_cpt_full), plot_R_full))

# Component rows
for (key in names(components)) {
  grid_grobs <- c(grid_grobs,
                  list(make_row_label(components[[key]]$label),
                       plots_F[[key]],
                       align_years(plots_SSB_cpt[[key]]),
                       plots_R[[key]]))
}

# Shared x-axis label plus a small legend for the reference lines
legend_grob <- grobTree(
  linesGrob(x = unit(c(0.38, 0.41), "npc"), y = unit(0.5, "npc"),
            gp = gpar(col = "#56B4E9", lwd = 2)),
  textGrob(expression(F[MSY]), x = unit(0.415, "npc"), y = unit(0.5, "npc"),
           just = "left", gp = gpar(fontsize = 10)),
  linesGrob(x = unit(c(0.49, 0.52), "npc"), y = unit(0.5, "npc"),
            gp = gpar(col = "darkorange", lwd = 2)),
  textGrob(expression("MSY B"[trigger]), x = unit(0.525, "npc"), y = unit(0.5, "npc"),
           just = "left", gp = gpar(fontsize = 10)))

grid_plot <- grid.arrange(
  grobs = grid_grobs,
  nrow = 6, ncol = 4,
  widths = c(0.15, 1, 1, 1),
  heights = c(0.1, 1, 1, 1, 1, 1),
  bottom = arrangeGrob(textGrob("Year", gp = gpar(fontsize = 12)),
                       legend_grob,
                       nrow = 2, heights = unit(c(1.2, 1.2), "lines")))
```

**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "cpt_grid.png"),
       plot = grid_plot,
       bg = "white",
       width = 12,
       height = 10,
       dpi = 300)
```



# Hysteresis
**Analysis:**
The SSB-F cross-correlation is examined per component; the resulting lag choices
are stored in the component configuration (SHOR 1, Buchan 3, Banks 5, Downs 1).
`breakpoint_analysis()` computes the SSB lag internally; a lagged SSB column is
also added for the hysteresis plots.

**Numerical note:** `breakpoint_analysis()` uses `strucchange::breakpoints()`
(via `opt_bpts()`), the same underlying method and helper as
`srr_breakpoint_analysis()` - a different algorithm from `bcp()`'s MCMC, and
not affected by that numerical underflow bug (verified: rescaling SSB/F by
several orders of magnitude does not change any breakpoint result here, unlike
the BCP case). `opt_bpts()` did have a real bug of its own though: when a
BIC-vs-breaks curve has more than one interior local minimum, it returned them
in scan order rather than ranked by fit, and callers always use the first
element - so a shallow, worse-fitting local minimum at a low break count could
silently be returned instead of a deeper, better-fitting one at a higher break
count. This was caught on Downs' hysteresis specifically: its BIC curve has
local minima at both 2 and 4 breaks, with 4 the true (lower-BIC) global
minimum, but the unfixed function returned 2. Checked against every other
hysteresis and SRR analysis in this project (all components and the full
stock) - none of the others have more than one interior local minimum, so this
was the only case actually affected; `opt_bpts()` has been fixed to rank
multiple local minima by BIC regardless.

```{r}
for (key in names(components)) {
  ccf(Herring[[key]]$SSB_component, Herring[[key]]$F, type = "correlation",
      main = paste("SSB x F -", components[[key]]$label))
}

results_brkpt <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  Herring[[key]]$SSB_lag <- dplyr::lag(Herring[[key]]$SSB_component, cfg$lag)
  results_brkpt[[key]] <- breakpoint_analysis(Herring[[key]], "SSB_component", "F", "year",
                                              lag_years = cfg$lag,
                                              region_name = cfg$label)
}
```


**Results:**
```
Shetland / Orkney (lag 1)
  Optimal number of breaks: 5
  Best breakpoint F values: 0.28, 0.29, 0.189, 0.29, 0.22
  Best breakpoint years:    1983, 1990, 1997, 2005, 2015

Buchan (lag 3)
  Optimal number of breaks: 3
  Best breakpoint F values: 0.29, 0.189, 0.184
  Best breakpoint years:    1990, 1997, 2012

Banks (lag 5)
  Optimal number of breaks: 2
  Best breakpoint F values: 0.39, 0.29
  Best breakpoint years:    1988, 2005

Downs (lag 1)
  Optimal number of breaks: 4
  Best breakpoint F values: 0.41, 0.171, 0.073, 0.24
  Best breakpoint years:    1987, 2002, 2009, 2016
```


**Plot:**

```{r}
plots_hysteresis <- list()
for (key in names(components)) {
  plots_hysteresis[[key]] <- plot_hysteresis(Herring[[key]],
                                             results_brkpt[[key]]$break_years,
                                             components[[key]]$label,
                                             show_msy_btrigger = FALSE)
  print(plots_hysteresis[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "hysteresis_components.png"),
       plot = grid.arrange(grobs = lapply(plots_hysteresis, strip_axes),
                           ncol = 2,
                           bottom = "F",
                           left = "SSB (million t)"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```



# SRR Non-Stationary
**Analysis:**
For each component: compare SRR models, validate with LOOCV and extract the
strucchange SRR breakpoints.

In the comparison tables, RMSE and R² are computed on the response scale for all
models, and AIC is only reported for models fitted on the raw response.
strucchange has the lowest in-sample RMSE for every component.

```{r}
source("R/model_comp_functions.R")

models_SRR <- list()
rmse_mean <- list()
results_SRR <- list()

for (key in names(components)) {
  cfg <- components[[key]]

  #### Compare different models
  models_SRR[[key]] <- run_srr_analysis(
    data_for_srr = Herring[[key]],
    recruit_col = "Recruitment",
    ssb_col = "SSB_component")

  print(models_SRR[[key]]$comparison_table)

  #### Prep data for LOOCV function & run loocv for all models
  dat <- Herring[[key]] %>%
    mutate(R_log = log(Recruitment),
           SSB_log = log(SSB_component)) %>%
    dplyr::select(year, Recruitment, SSB_component, R_log, SSB_log) %>%
    na.omit() %>%
    setNames(c("Year", "R", "SSB", "R_log", "SSB_log")) %>%
    as.data.frame()

  rmse_mean[[key]] <- cross_valid(dataset = dat)

  #### Extract SRR breakpoints
  results_SRR[[key]] <- srr_breakpoint_analysis(Herring[[key]], "SSB_component",
                                                "Recruitment", "year", cfg$label)
}
```


**Results:**

LOOCV note: `cross_valid()` also reports how many of the folds produced a
test prediction per model (`n_test_folds`), a mean over few folds is not
comparable to models evaluated on all folds. The values below are from a
verified re-run.

**Shetland / Orkney** (53 folds, all models evaluated on all folds)
```
LOOCV   linear    beverton    ricker      segmented   seg.log     segm.glm    strucchange
TEST:   6130646   5993134     6077639     5862612     5567516     5993610     8907639
TRAIN:  7308756   8173777     8045893     6787493     7062180     7092471     4388715

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 150184.2, 511033.1, 1223559.3
Best breakpoint years:      1980, 1987, 2001
```

LOOCV: strucchange has the lowest train RMSE but the highest test RMSE, fits
the observed series closely while predicting left-out years poorly.

**Buchan** (53 folds; beverton evaluated on 52)
```
LOOCV   linear    beverton    ricker    segmented   seg.log     segm.glm    strucchange
TEST:   3975684   3596876     3692709   3964665     3466497     4275538     8939867
TRAIN:  5980044   6376121     6353865   5920518     6301303     6093582     2864856

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 149083.9, 1278132.6, 125887.2
Best breakpoint years:      1983, 1991, 2001
```

LOOCV: same pattern as Shetland / Orkney, strucchange fits the series best
(lowest train RMSE) but predicts left-out years worst (highest test RMSE).

**Banks** (53 folds, all models evaluated on all folds)
```
LOOCV   linear    beverton    ricker    segmented   seg.log     segm.glm    strucchange
TEST:   2303908   1979821     1963756   2510689     2029915     2362265     4314485
TRAIN:  3262373   3230360     3182808   3053152     3160261     3067352     1405691

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 38669.1, 311167.9, 309162.5
Best breakpoint years:      1980, 1987, 2001
```

LOOCV: same pattern as Shetland / Orkney, strucchange fits the series best
(lowest train RMSE) but predicts left-out years worst (highest test RMSE).

**Downs** (53 folds; beverton evaluated on only 1 fold, strucchange on only 5 —
their test means are **not** comparable and should not be interpreted)
```
LOOCV   linear    beverton*   ricker      segmented   seg.log     segm.glm    strucchange*
TEST:   1552443   (13503)     1466005     1285054     1274572     1290043     (3032227)
TRAIN:  1862654   1853802     2031077     1695986     1768639     1690839     1503568

SRR breakpoints: optimal number of breaks: 0 (relationship treated as continuous)
```

LOOCV: among the models evaluated on all folds, the segmented variants predict
held-out years best and the linear model worst.

**Downs no longer has any SRR breakpoint support.** With the updated ICES data
the full-series BIC minimum for `strucchange` is 0 breaks outright — there is no
longer even a weakly-supported local optimum to report (previously an interior
BIC local minimum surfaced 4 breakpoints via the `opt_bpts` rule; that local
optimum no longer exists). `srr_breakpoint_analysis()` now returns `break_years
= numeric(0)` for Downs and `plot_SRR()` draws it as a single continuous
relationship (subtitle "continuous (no breakpoints)") instead of a
strucchange-segmented one.

**Failed Folds (Buchan, Downs), and what that means:**
* *Beverton-Holt (1/53 folds usable for Downs):* the log-scale nls fit aborts
  with non-finite values for 52 of 53 training subsets, the saturating BH curve
  shape is essentially incompatible with the increasing Downs SRR data, so the
  model cannot even be evaluated for almost any subset.
* *strucchange (5/53 folds usable for Downs):* for the large majority of
  training subsets the BIC-minimizing partition selects **zero** breakpoints;
  without breakpoints no piecewise prediction exists and the fold is skipped.
  Together with the absence of consensus changepoints in its SSB series (see
  changepoint section), this indicates that Downs shows no signs of abrupt
  regime shifts; its LOOCV favoring the smooth segmented models points to
  gradual nonlinear change instead of a discrete shift.

**Overall:** across Shetland/Orkney, Buchan and Banks, strucchange describes the
observed regimes most closely (lowest train RMSE) while its held-out
predictions are mostly the weakest, consistent with it capturing regime
structure rather than a single predictive stock-recruitment relationship. Downs
is the exception: no regime structure is supported at all, and the smooth
continuous models both fit and predict better than any breakpoint model.


**Plot:**

```{r}
plots_SRR <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  no_breaks <- results_SRR[[key]]$optimal_breaks == 0
  plots_SRR[[key]] <- plot_SRR(Herring[[key]],
                               break_years = results_SRR[[key]]$break_years,
                               title_stock = cfg$label,
                               used_model = if (no_breaks) "continuous (no breakpoints)" else "strucchange",
                               ssb_col = "SSB_component",
                               show_Blim = FALSE,
                               nudge_params = cfg$srr_nudge)
  print(plots_SRR[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "SRR_components.png"),
       plot = grid.arrange(grobs = lapply(plots_SRR, strip_axes),
                           ncol = 2,
                           bottom = "SSB (million t)",
                           left = "R (billions)"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```