---
title: "data cleanup"
author: "finn krauss"
date: "2025-05-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = here::here())
```



# Load packages

```{r}
# installs packages if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(tidyverse, here)
```



## Load and clean files
**LAI Data Import:**

```{r}
## Read LAI .txt data, rename its columns & convert from century break to full year
LAI_full <- read.table("data raw/NSAS LAI.txt", header = TRUE, strip.white = TRUE) %>%
  rename(component = Area,
         year = Year,
         LAI_unit = LAIUnit,
         fortnight = Fortnight,
         LAI_9mm = L..9) %>%
  mutate(year = ifelse(year >= 100, 2000 + (year - 100),1900 + year),
         component = case_when(component == "Or/Sh" ~ "Shetland_Orkney",
                               component == "CNS" ~ "Banks",
                               component == "SNS" ~ "Downs",
                               TRUE ~ component),
         component = as.factor(component),
         LAI_unit = as.factor(LAI_unit),
         fortnight = as.factor(fortnight))

## Read LAI_components .csv, convert it into longformat & rename columns for continuity
LAI_component <- read_csv("data raw/NSAS LAI Components 2025.csv") %>%
  pivot_longer(cols = -c(1), names_to = "year", values_to = "LAI_perc.") %>%
  rename(component = ...1) %>%
  mutate(component = case_when(component == "LAI-ORSH" ~ "Shetland_Orkney",
                               component == "LAI-BUN" ~ "Buchan",
                               component == "LAI-CNS" ~ "Banks",
                               component == "LAI-SNS" ~ "Downs"),
         component = as.factor(component),
         year = as.numeric(year),
         LAI_perc. = LAI_perc. * 100)

## Read ICES 2026 .txt data (SSB & assessment combined) & rename columns for continuity
Herring_full <- read.csv("data raw/NSAS ICES data 2026.txt") %>%
  rename(l_bnd = SSB_low, u_bnd = SSB_high,
         Recruitment_high = recruitment_high,
         Recruitment_low = recruitment_low)
```


Variables: SSB = SSB in Tonnes, l_bnd = lower limit of 95% CI, u_bnd = upper
limit of 95% CI, F = fishing mortality, Recruitment = recruitment in numbers.



## Split Components

The SSB data (with its 95% bounds) and the recruitment are split by each component's LAI percentage.

The split pairs year-y shares with year-y values for SSB **and** recruitment, no larval-to-recruitment lag.

```{r}
Herring_LAI <- Herring_full %>%
  left_join(LAI_component, by = "year") %>%
  mutate(SSB_component = SSB * (LAI_perc. / 100),
         l_bnd_component = l_bnd * (LAI_perc. / 100),
         u_bnd_component = u_bnd * (LAI_perc. / 100),
         Recruitment = Recruitment * (LAI_perc. / 100),
         Recruitment_high = Recruitment_high * (LAI_perc. / 100),
         Recruitment_low = Recruitment_low * (LAI_perc. / 100)) %>%
  dplyr::select(year, component, LAI_perc., SSB_component, l_bnd_component, u_bnd_component, F, F_high, F_low, Recruitment, Recruitment_high, Recruitment_low) %>%
  drop_na()
```



## Visual investigation

```{r}
Herring_LAI$component <- factor(Herring_LAI$component, levels = c("Shetland_Orkney", "Buchan", "Banks", "Downs"))
LAI_component$component <- factor(LAI_component$component, levels = c("Shetland_Orkney", "Buchan", "Banks", "Downs"))

## Plot timeseries of LAI by component
### Add LAI by year
LAI_aggregated <- LAI_full %>%
  group_by(year, component) %>%
  summarise(LAI_9mm_sum = sum(LAI_9mm),
    .groups = 'drop') %>%
  mutate(component = factor(component, levels = c("Shetland_Orkney", "Buchan", "Banks", "Downs")))

### Plot aggregated LAI
plot_LAI_byyear <- ggplot(LAI_aggregated, aes(x = year,
                    y = LAI_9mm_sum,
                    color = component,
                    shape = component)) +
  geom_point(size = 1.5, alpha = 0.8) +
  geom_line(linewidth = 0.8, alpha = 0.8) +
  scale_color_viridis_d(option = "mako", begin = 0.1, end = 0.8) +
  scale_shape_manual(values = c("Shetland_Orkney" = 16, "Buchan" = 17, "Banks" = 15, "Downs" = 18)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(title = NULL,
       x = "Year",
       y = "LAI",
       color = "Component",
       shape = "Component") +
  guides(color = guide_legend(), shape = guide_legend()) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot_LAI_byyear

## Plot stacked bar chart for components percentage of LAI
plot_LAI_percent <- ggplot(LAI_component, aes(x = as.factor(year),
                              y = LAI_perc., fill = component)) +
  geom_bar(stat = "identity", position = "stack") + # Create stacked bars
  scale_fill_viridis_d(option = "mako", begin = 0.1, end = 0.8) +  # Use mako palette with capped range
  scale_x_discrete(breaks = function(x) x[seq(1, length(x), by = 2)]) +  # Show every second year
  labs(title = NULL,
       x = "Year",
       y = "LAI Percentage (%)",
       fill = "Component") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot_LAI_percent

### SSB component overlay
plot_SSB_component <- ggplot(Herring_LAI, aes(x = year, y = SSB_component / 1e6, color = component, fill = component)) +
  geom_ribbon(aes(ymin = l_bnd_component / 1e6,
                  ymax = u_bnd_component / 1e6), alpha = 0.3, color = NA) +
  geom_line(linewidth = 0.8) +
  scale_color_viridis_d(option = "mako", begin = 0.1, end = 0.8) +
  scale_fill_viridis_d(option = "mako", begin = 0.1, end = 0.8) +
  labs(title = NULL,
       x = "Year",
       y = "SSB (million t)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot_SSB_component

### R component overlay
plot_R_component <- ggplot(Herring_LAI, aes(x = year, y = Recruitment / 1e6, color = component, fill = component)) +
  geom_line(linewidth = 0.8) +
  scale_color_viridis_d(option = "mako", begin = 0.1, end = 0.8) +
  scale_fill_viridis_d(option = "mako", begin = 0.1, end = 0.8) +
  labs(title = NULL,
       x = "Year",
       y = "R (billions)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        axis.title.x = element_text(margin = margin(t = 10)),
        axis.title.y = element_text(margin = margin(r = 10)))
plot_R_component
```



## Save data and plots
**Save Rdata:**

```{r}
### LAI data
save(LAI_full, file = file.path("data/", "LAI_full.Rdata"))
save(LAI_aggregated, file = file.path("data/", "LAI_aggregated.Rdata"))
save(Herring_full, file = file.path("data/", "Herring_full.Rdata"))
save(Herring_LAI, file = file.path("data/", "Herring_LAI.Rdata"))
```


**Save Plots:**

```{r}
### LAI plots
ggsave(filename = file.path("plots/", "LAI_by_year.png"),
       plot = plot_LAI_byyear,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)

ggsave(filename = file.path("plots/", "LAI_percent.png"),
       plot = plot_LAI_percent,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)

ggsave(filename = file.path("plots/", "SSB_component.png"),
       plot = plot_SSB_component,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)

ggsave(filename = file.path("plots/", "R_component.png"),
       plot = plot_R_component,
       bg = "white",
       width = 7,
       height = 5,
       dpi = 300)
```