---
title: "data_analysis_component"
author: "finn krauss"
date: "2025-05-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath(".."))
```



## Load Packages

```{r}
# installs packages if not already installed
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")

pacman::p_load(bcp, changepoint, ggrepel,
  ggplot2, tidyverse, FSA, strucchange,
  MASS, segmented, gridExtra)
```



## Import data & component configuration
The `components` list holds the per-component configuration:

* **name** - component name in `Herring_LAI`
* **label** - label used in plot titles and printed output
* **lag** - SSB-F time lag in years, chosen from the CCF plots below
* **srr_nudge** - label nudges for the SRR plots
  (first year, breakpoint, last year, more breakpoints)

In all changepoint plots the CI ribbons are colored by the **full-stock phases**
(`full_changepoints`: pre-1966 #201124, 1966–2000 #3e478d, post-2000 #318ca5),
so the full-stock changepoints are visible in every component panel and it is
directly comparable how the components' own shifts align with them. Each
component's own consensus changepoints are drawn as thin black lines on top.
Buchan and Downs have no consensus changepoints.

```{r}
load("data/Herring_LAI.Rdata")

source("R/functions.R")

### Full-stock phases used to color the CI ribbons in all changepoint plots
full_changepoints <- c(1966, 2000)
full_ribbon <- c("#201124", "#3e478d", "#318ca5")
full_vline <- c("black", "black")

components <- list(
  SHOR = list(
    name = "Shetland_Orkney",
    label = "Shetland / Orkney",
    lag = 1,
    srr_nudge = list(
      list(nudge_y = 9, nudge_x = -0.15),
      list(nudge_y = 0, nudge_x = -0.2),
      list(nudge_y = 20, nudge_x = 0),
      list(nudge_y = -0, nudge_x = 0.2))),
  Buchan = list(
    name = "Buchan",
    label = "Buchan",
    lag = 3,
    srr_nudge = list(
      list(nudge_y = 5, nudge_x = 0),
      list(nudge_y = 5, nudge_x = 0),
      list(nudge_y = 25, nudge_x = 0),
      list(nudge_y = 10, nudge_x = -0.1))),
  Banks = list(
    name = "Banks",
    label = "Banks",
    lag = 5,
    srr_nudge = list(
      list(nudge_y = 10, nudge_x = 0),
      list(nudge_y = 8, nudge_x = 0),
      list(nudge_y = 15, nudge_x = 0),
      list(nudge_y = 0, nudge_x = 0.05))),
  Downs = list(
    name = "Downs",
    label = "Downs",
    lag = 1,
    srr_nudge = list(
      list(nudge_y = 7, nudge_x = 0),
      list(nudge_y = 4, nudge_x = 0),
      list(nudge_y = 7, nudge_x = 0),
      list(nudge_y = 5, nudge_x = 0))))

### Split data into components
Herring <- lapply(components, function(cfg) {
  Herring_LAI[Herring_LAI$component == cfg$name, ]
})

rm(Herring_LAI)

### Helper to strip axis titles when stacking plots into grids
strip_axes <- function(p) {
  p + theme(axis.title.x = element_blank(), axis.title.y = element_blank())
}
```



# Change-Points
**Analysis:**
`bcp()` estimates posterior probabilities by MCMC. The thesis run used the bcp
default of 500 iterations without a seed, which is not converged: re-running
changed the changepoint sets. `changepoint_analysis()` now uses a fixed seed and
5000 iterations; with the converged chain the results are stable across seeds.

```{r}
results_cpt <- list()
for (key in names(components)) {
  results_cpt[[key]] <- changepoint_analysis(Herring[[key]],
                                             region_name = components[[key]]$label,
                                             year_col = "year",
                                             Q = 6,
                                             bcp_threshold = 0.7,
                                             ssb_col = "SSB_component")
}
```


**Results:**
**Shetland / Orkney** (unchanged from thesis)
```
CPT changepoint years:  1983, 1990, 1994, 1996, 2005, 2014
BCP changepoint years:  1983, 1990, 1994, 2002, 2005, 2014
Consensus changepoints (± 1 yr, min 5 yr apart): 1983, 1990, 2005, 2014
```

**Banks** (unchanged from thesis)
```
CPT changepoint years:  1982, 1990, 2000, 2005, 2009, 2019
BCP changepoint years:  1983, 1990, 2000, 2001, 2005
Consensus changepoints (± 1 yr, min 5 yr apart): 1982, 1990, 2000, 2005
```

**Buchan & Downs** (changed from thesis)
With the converged chain the BCP posterior changepoint probability is ~0 for
every year in both components, no year reaches the 0.7 threshold, so BCP
detects **no** changepoints.

No consensus changepoints exist for Buchan and. (For
reference, CPT alone suggests Buchan: 1986, 1989, 1993, 2009, 2011, 2018 and
Downs: 1987, 1999, 2001, 2011, 2015, 2020, but without BCP support these do not
qualify as consensus changepoints.)

*Interpretation:* for Downs this converges with the SRR analysis (see below),
where the breakpoint model is also only weakly supported, Downs shows weaker
signs of abrupt regime shifts than the other components. For Buchan the evidence
is mixed: no SSB changepoints, but supported SRR breakpoint structure.

**Plot:**

```{r}
plots_SSB_cpt <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  cpts <- results_cpt[[key]]$consensus$changepoint_years
  plots_SSB_cpt[[key]] <- plot_SSB_cpt(data = Herring[[key]],
                changepoints = cpts,
                ribbon_changepoints = full_changepoints,
                ribbon_colors = full_ribbon,
                vline_colors = rep("black", length(cpts)),
                component = cfg$label,
                SSB_column = "SSB_component",
                l_bnd_column = "l_bnd_component",
                u_bnd_column = "u_bnd_component",
                show_hlines = FALSE)
  print(plots_SSB_cpt[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "SSB_cpt_components.png"),
       plot = grid.arrange(grobs = lapply(plots_SSB_cpt, strip_axes),
                           ncol = 2,
                           bottom = "Year",
                           left = "SSB in million t"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```

**Plot Grid:**

```{r}
load("data/Herring_full.Rdata")

### Align the SSB year axis with the F and R columns (fixed 1947-2024 range) and
### drop the per-panel x-axis title (one shared "Year" label sits below the grid)
align_years <- function(p) {
  suppressWarnings(p + xlim(1947, 2024) + theme(axis.title.x = element_blank()))
}

plot_SSB_cpt_full <- plot_SSB_cpt(data = Herring_full,
                                  changepoints = full_changepoints,
                                  ribbon_colors = full_ribbon,
                                  vline_colors = full_vline,
                                  component = "Full Stock",
                                  SSB_column = "SSB",
                                  l_bnd_column = "l_bnd",
                                  u_bnd_column = "u_bnd",
                                  show_hlines = TRUE)

plot_F_full <- plot_F_cpt(data = Herring_full,
                          changepoints = full_changepoints,
                          ribbon_colors = full_ribbon,
                          vline_colors = full_vline,
                          component = "Full Stock",
                          show_hlines = TRUE)

plot_R_full <- plot_Recruitment_cpt(data = Herring_full,
                                    changepoints = full_changepoints,
                                    ribbon_colors = full_ribbon,
                                    vline_colors = full_vline,
                                    component = "Full Stock")

# Component F and R plots (ribbons colored by the full-stock phases)
plots_F <- list()
plots_R <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  cpts <- results_cpt[[key]]$consensus$changepoint_years

  plots_F[[key]] <- plot_F_cpt(data = Herring[[key]],
                               changepoints = cpts,
                               ribbon_changepoints = full_changepoints,
                               ribbon_colors = full_ribbon,
                               vline_colors = rep("black", length(cpts)),
                               component = cfg$label,
                               show_hlines = TRUE)

  plots_R[[key]] <- plot_Recruitment_cpt(data = Herring[[key]],
                                         changepoints = cpts,
                                         ribbon_changepoints = full_changepoints,
                                         ribbon_colors = full_ribbon,
                                         vline_colors = rep("black", length(cpts)),
                                         component = cfg$label)
}

# Assemble the grid with row labels and column headers
library(grid)

make_row_label <- function(label) {
  textGrob(label, rot = 90, gp = gpar(fontsize = 12, fontface = "bold"))
}
make_col_header <- function(label) {
  textGrob(label, gp = gpar(fontsize = 12, fontface = "bold"))
}

grid_grobs <- c(
  # Header row
  list(nullGrob(),
       make_col_header("Fishing Mortality"),
       make_col_header("SSB with Changepoints"),
       make_col_header("Recruitment")),
  # Full stock row
  list(make_row_label("Full Stock"), plot_F_full,
       align_years(plot_SSB_cpt_full), plot_R_full))

# Component rows
for (key in names(components)) {
  grid_grobs <- c(grid_grobs,
                  list(make_row_label(components[[key]]$label),
                       plots_F[[key]],
                       align_years(plots_SSB_cpt[[key]]),
                       plots_R[[key]]))
}

# Shared x-axis label plus a small legend for the reference lines
legend_grob <- grobTree(
  linesGrob(x = unit(c(0.38, 0.41), "npc"), y = unit(0.5, "npc"),
            gp = gpar(col = "#56B4E9", lwd = 2)),
  textGrob(expression(F[MSY]), x = unit(0.415, "npc"), y = unit(0.5, "npc"),
           just = "left", gp = gpar(fontsize = 10)),
  linesGrob(x = unit(c(0.49, 0.52), "npc"), y = unit(0.5, "npc"),
            gp = gpar(col = "darkorange", lwd = 2)),
  textGrob(expression("MSY B"[trigger]), x = unit(0.525, "npc"), y = unit(0.5, "npc"),
           just = "left", gp = gpar(fontsize = 10)))

grid_plot <- grid.arrange(
  grobs = grid_grobs,
  nrow = 6, ncol = 4,
  widths = c(0.15, 1, 1, 1),
  heights = c(0.1, 1, 1, 1, 1, 1),
  bottom = arrangeGrob(textGrob("Year", gp = gpar(fontsize = 12)),
                       legend_grob,
                       nrow = 2, heights = unit(c(1.2, 1.2), "lines")))
```

**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "cpt_grid.png"),
       plot = grid_plot,
       bg = "white",
       width = 12,
       height = 10,
       dpi = 300)
```



# Hysteresis
**Analysis:**
The SSB-F cross-correlation is examined per component; the resulting lag choices
are stored in the component configuration (SHOR 1, Buchan 3, Banks 5, Downs 1).
`breakpoint_analysis()` computes the SSB lag internally; a lagged SSB column is
also added for the hysteresis plots.

```{r}
for (key in names(components)) {
  ccf(Herring[[key]]$SSB_component, Herring[[key]]$F, type = "correlation",
      main = paste("SSB x F -", components[[key]]$label))
}

results_brkpt <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  Herring[[key]]$SSB_lag <- dplyr::lag(Herring[[key]]$SSB_component, cfg$lag)
  results_brkpt[[key]] <- breakpoint_analysis(Herring[[key]], "SSB_component", "F", "year",
                                              lag_years = cfg$lag,
                                              region_name = cfg$label)
}
```


**Results:**
```
Shetland / Orkney (lag 1)
  Optimal number of breaks: 5
  Best breakpoint F values: 0.28, 0.29, 0.189, 0.29, 0.22
  Best breakpoint years:    1983, 1990, 1997, 2005, 2015

Buchan (lag 3)
  Optimal number of breaks: 3
  Best breakpoint F values: 0.29, 0.189, 0.184
  Best breakpoint years:    1990, 1997, 2012

Banks (lag 5)
  Optimal number of breaks: 2
  Best breakpoint F values: 0.39, 0.29
  Best breakpoint years:    1988, 2005

Downs (lag 1)
  Optimal number of breaks: 4
  Best breakpoint F values: 0.41, 0.171, 0.073, 0.24
  Best breakpoint years:    1987, 2002, 2009, 2016
```


**Plot:**

```{r}
plots_hysteresis <- list()
for (key in names(components)) {
  plots_hysteresis[[key]] <- plot_hysteresis(Herring[[key]],
                                             results_brkpt[[key]]$break_years,
                                             components[[key]]$label,
                                             show_msy_btrigger = FALSE)
  print(plots_hysteresis[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "hysteresis_components.png"),
       plot = grid.arrange(grobs = lapply(plots_hysteresis, strip_axes),
                           ncol = 2,
                           bottom = "F",
                           left = "SSB in million t"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```



# SRR Non-Stationary
**Analysis:**
For each component: compare SRR models, validate with LOOCV and extract the
strucchange SRR breakpoints.

In the comparison tables, RMSE and R² are computed on the response scale for all
models, and AIC is only reported for models fitted on the raw response.
strucchange has the lowest in-sample RMSE for every component.

```{r}
source("R/model_comp_functions.R")

models_SRR <- list()
rmse_mean <- list()
results_SRR <- list()

for (key in names(components)) {
  cfg <- components[[key]]

  #### Compare different models
  models_SRR[[key]] <- run_srr_analysis(
    data_for_srr = Herring[[key]],
    recruit_col = "Recruitment",
    ssb_col = "SSB_component")

  print(models_SRR[[key]]$comparison_table)

  #### Prep data for LOOCV function & run loocv for all models
  dat <- Herring[[key]] %>%
    mutate(R_log = log(Recruitment),
           SSB_log = log(SSB_component)) %>%
    dplyr::select(year, Recruitment, SSB_component, R_log, SSB_log) %>%
    na.omit() %>%
    setNames(c("Year", "R", "SSB", "R_log", "SSB_log")) %>%
    as.data.frame()

  rmse_mean[[key]] <- cross_valid(dataset = dat)

  #### Extract SRR breakpoints
  results_SRR[[key]] <- srr_breakpoint_analysis(Herring[[key]], "SSB_component",
                                                "Recruitment", "year", cfg$label)
}
```


**Results:**

LOOCV note: `cross_valid()` also reports how many of the folds produced a
test prediction per model (`n_test_folds`), a mean over few folds is not
comparable to models evaluated on all folds. The values below are from a
verified re-run.

**Shetland / Orkney** (52 folds, all models evaluated on all folds)
```
LOOCV   linear    beverton    ricker      segmented   seg.log     segm.glm    strucchange
TEST:   6032494   5917938     5996407     5906258     5705907     5919391     9093970
TRAIN:  7193145   7998380     7896592     6659682     6947653     6985388     4409284

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 148853.8, 505309.4, 1201916.7
Best breakpoint years:      1980, 1987, 2001
```

LOOCV: strucchange has the lowest train RMSE but the highest test RMSE, fits
the observed series closely while predicting left-out years poorly.

**Buchan** (52 folds; beverton evaluated on 51)
```
LOOCV   linear    beverton    ricker    segmented   seg.log     segm.glm    strucchange
TEST:   3896840   3509676     3594421   3800680     3380906     4106902     8445141
TRAIN:  5799156   6165427     6142351   5736075     6092541     5923863     2828094

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 147574.4, 1257071.9, 123660.5
Best breakpoint years:      1983, 1991, 2001
```

LOOCV: same pattern as Shetland / Orkney, strucchange fits the series best
(lowest train RMSE) but predicts left-out years worst (highest test RMSE).

**Banks** (52 folds, all models evaluated on all folds)
```
LOOCV   linear    beverton    ricker    segmented   seg.log     segm.glm    strucchange
TEST:   2298870   1970034     1954005   2399081     2038932     2370784     4418968
TRAIN:  3211518   3180706     3133516   2988736     3112284     3020506     1363380

SRR breakpoints: optimal number of breaks: 3
Best breakpoint SSB values: 38326.5, 307682.7, 303694
Best breakpoint years:      1980, 1987, 2001
```

LOOCV: same pattern as Shetland / Orkney, strucchange fits the series best
(lowest train RMSE) but predicts left-out years worst (highest test RMSE).

**Downs** (52 folds; beverton evaluated on only 1 fold, strucchange on only 7 —
their test means are **not** comparable and should not be interpreted)
```
LOOCV   linear    beverton*   ricker      segmented   seg.log     segm.glm    strucchange*
TEST:   1504907   (13872)     1413304     1245100     1228831     1235682     (2689827)
TRAIN:  1784159   1771427     1936653     1619286     1688248     1611886     1393370

SRR breakpoints: optimal number of breaks: 4
Best breakpoint SSB values: 17120.8, 213659.7, 565941.9, 435906.7
Best breakpoint years:      1980, 2000, 2007, 2014
```

LOOCV: among the models evaluated on all folds, the segmented variants predict
held-out years best and the linear model worst; strucchange again has the lowest
train RMSE.

**Failed Folds (Buchan, Downs), and what that means:**
* *Beverton-Holt (1/52 folds usable):* the log-scale nls fit aborts with
  non-finite values for 51 of 52 training subsets, the saturating BH curve
  shape is essentially incompatible with the increasing Downs SRR data, so the
  model cannot even be evaluated for almost any subset.
* *strucchange (7/52 folds usable):* for 45 of 52 training subsets the
  BIC-minimizing partition selects **zero** breakpoints; without breakpoints no
  piecewise prediction exists and the fold is skipped. The same holds for the
  full Downs series: the global BIC minimum is 0 breaks, and the 4 reported
  breakpoints correspond to an interior local minimum of the BIC curve
  (`opt_bpts` rule), the Downs breakpoint model is only weakly supported.
  Together with the absence of consensus changepoints in its SSB series (see
  changepoint section), this indicates that Downs shows weaker signs of abrupt
  regime shifts than the other components; its LOOCV favoring the smooth
  segmented models points to gradual nonlinear change instead.

**Overall:** across components, strucchange describes the observed regimes most
closely (lowest train RMSE throughout) while its held-out predictions are mostly
the weakest, consistent with it capturing regime structure rather than a single
predictive stock-recruitment relationship.


**Plot:**

```{r}
plots_SRR <- list()
for (key in names(components)) {
  cfg <- components[[key]]
  plots_SRR[[key]] <- plot_SRR(Herring[[key]],
                               break_years = results_SRR[[key]]$break_years,
                               title_stock = cfg$label,
                               used_model = "strucchange",
                               ssb_col = "SSB_component",
                               show_Blim = FALSE,
                               nudge_params = cfg$srr_nudge)
  print(plots_SRR[[key]])
}
```



**Save Plot:**

```{r}
ggsave(filename = file.path("plots/", "SRR_components.png"),
       plot = grid.arrange(grobs = lapply(plots_SRR, strip_axes),
                           ncol = 2,
                           bottom = "SSB in million t",
                           left = "R in billions"),
       bg = "white",
       width = 12,
       height = 9,
       dpi = 300)
```